Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-07 Dail_Mad_ext.tex
ss equations in intro

\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines,displaymath]{lineno}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}

@@

One of the simplest state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a deterministic population growth model
(e.g. exponential, logistic, etc \dots)~and a stochastic model allowing
for random process variation, which we denote $\grave{o}_t$. As an
example, under density-independent exponential population growth, the
model could be written as:
\begin{equation}
%  \label{eq:ss1}
  \left.
  \begin{aligned}
    \mu_t &= N_{t-1}\exp(r) \label{eq:ss1a} \\ %\label{eq:ss1b}
    N_t &\sim \text{N}(\mu_t, \grave{o}_t) \\ %\label{eq:ss1c}
    X_t &\sim \text{N}(N_t, \zeta_t)
  \end{aligned} \;
  \right\} \; \text{for} \; t=2,\hdots,T
\end{equation}
%\begin{subequations}
%  \label{eq:ss1}
%  \left.
%  \begin{gather}
%    \mu_t = N_{t-1}\exp(r) \label{eq:ss1a} \\ \label{eq:ss1b}
%    N_t \sim \text{Normal}(\mu_t, \grave{o}_t) \\ \label{eq:ss1c}
%    X_t \sim \text{Normal}(N_t, \zeta_t)
%  \end{gather}
%  \right\} \text{for}
%\end{subequations}
where $r$ is the intrinsic rate of increase. Note that no model is
presented for the initial state at $t=1$---this issue will be
discussed subsequently. Note also that the exponential
growth model shown in Eq.~\ref{eq:ss1a} can be easily replaced with
any population growth model, including density-depedent models. This
flexibility is an appealing feature of state-space models. One feature
that is common to classical state-space models is that the
error terms are assumed to be i.i.d random effects following Gaussian
(Normal) distributions:
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. The Gaussian assumption allows for parameter
estimation using the Kalman filter \citep{dennis_etal:2006}, which is
much more computationally efficient than estimation methods when
random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
are the most widespread approach for modeling population dynamics
using time-series data, several problems are evident. Namely, (1)
time-series data and the underlying abundance parameters of interest
are typically integer-valued, raising concerns about the use of
Gaussian distribution shown in Eq~\ref{eq:ss1b}, (2) the model for
observatoin error (Eq.~\ref{eq:ss1c}) has little biological basis, (3)
spatial variation in abundance is not allowed, and (4) some of the
paramters of the model are not estimable. We briefly discuss each of
these points before describing a general approach to resolving these
shortcomings.

The fact that both abundance ($N_t$) and the observed data ($X_t$) are
typically non-negative integers suggests that the
Gaussian assumption in Eq.~\ref{eq:ss1b} is not justified.
%is clearly a problem with this formulation because the normally distributed
%random effects violate this constraint.
Even so, efforts are often made to adhere to the
Gaussian assumption by tranforming abundance to
density by replacing $N_t$ with $Y_t = \log(N_t/A)$ where $A$ is the
area surveyed. However, this transformation is problematic since
abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with some small number, although the
effect of this is rarely discussed.

@@

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Specifically, the Markovian
nature of the model implies the following likelihood
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
    [N_1&] \\ \label{eq:ss1b}
    [N_t&|N_{t-1}] &\text{for} \; t=2,\hdots,T \\ \label{eq:ss1c}
    [X_t&|N_t]  &\text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
%\begin{equation}
%\mathcal{L}(\Theta|\{X_t\}) = [N_1|\Theta]\prod_{t=2}^T
%[N_t|N_{t-1},\Theta]
%\label{eq:L1}
%\end{equation}
The first term is the probability density for the first observation in
the time series, $X_1$, yet it is intuitively not possible to estimate the
parameter(s) of this distribution using only a single
observation.
%This problem has resulted in several papers with
%interesting names like ``multi-modal likelihoods and ridges etc...''
Two options have been proposed handling the identifiability
problem. First, one can assume that the first observation in the time
series is made without error. Second, one can assume that the population is at equilibrium
so that that the x can be replaced with the equilibrium
distribution. However, assuming equilibrium defeats the objective of
many studies of population dynamics, namely determining if and why a
population is at equilibrium.

>>>>

\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines,displaymath]{lineno}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}

@@

One of the simplest state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a deterministic population growth model
(e.g. exponential, logistic, etc \dots)~and a stochastic model allowing
for random process variation, which we denote $\grave{o}_t$. As an
example, under density-independent exponential population growth, the
model could be written as:
% \begin{equation}
%   \left.
%   \begin{aligned}
%     \mu_t &= N_{t-1}\exp(r) \\ %\label{eq:ss1b}
%     N_t &= \mu_t + \grave{o}_t \\ %\text{N}(\mu_t, \grave{o}_t) \\ %\label{eq:ss1c}
%     X_t &= N_t + \zeta_t %\text{N}(N_t, \zeta_t)
%   \end{aligned} \;
%   \right\} \; \text{for} \; t=2,\hdots,T
%   \label{eq:ss1}
% \end{equation}
\begin{subequations}
  \begin{align}
%    \mu_t &= N_{t-1}\exp(r) \quad \text{for} \; t=2,\hdots,T
    N_1 &= X_1 \label{ss1a} \\
    N_t &= \mu_t + \grave{o}_{t-1} \quad \text{for} \; t=2,\hdots,T \label{eq:ss1b} \\
    X_t &= N_t + \zeta_t \quad\;\;\; \text{for} \; t=1,\hdots,T \label{eq:ss1c}
  \end{align}
  \label{eq:ss1}
\end{subequations}
where $\mu_t$ is a deterministic function describing population
growth. For instance, under exponential growth, $\mu_t =
N_{t-1}e^{r}$ where $r$ is the intrinsic rate of
increase. Alternatively, one might consider density-dependent models,
and.  This
flexibility is an appealing feature of state-space models.
Note that no model is
presented for the initial state at $t=1$; we will return to this issue
subsequently. Note also that the exponential
growth model shown in Eq.~\ref{eq:ss1} can be easily replaced with
any population growth model, including density-depedent models. One feature
that is common to most classical state-space models is that the
error terms are assumed to be i.i.d random effects following Gaussian
(Normal) distributions:
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. The Gaussian assumption allows for parameter
estimation using the Kalman filter \citep{dennis_etal:2006}, which is
much more computationally efficient than estimation methods when
random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
are the most widespread approach for modeling population dynamics
using time-series data, several problems are evident. Namely, (1)
time-series data and the underlying abundance parameters of interest
are typically integer-valued, raising concerns about the use of
Gaussian distribution for the random effects, (2) the model for
observatoin error has little biological basis, (3)
spatial variation in abundance is not allowed, and (4) some of the
paramters of the model are not estimable. We briefly discuss each of
these points before describing a general approach to resolving these
shortcomings.

The fact that both abundance ($N_t$) and the observed data ($X_t$) are
typically non-negative integers suggests that it is inappropriate to
model process variation and observation error using Gaussian
distributions. Nonetheless, efforts have been made to adhere to the
Gaussian assumptions by tranforming abundance to
density by replacing $N_t$ with $Y_t = \log(N_t/A)$ where $A$ is the
area surveyed. However, this transformation is problematic since
abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with some small number, although the
effect of this is rarely discussed.

@@

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$ or
$X_1$; however, such distributions are needed to fully specify the
model. Indeed, a fully-specified state-space model requires the
specification of three probability distributions, which we denote
using bracket notation:
\begin{subequations}
  \label{eq:ss2}
  \begin{align}
    [N_1&|\bm{\theta}] \\ \label{eq:ss2b}
    [N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\ \label{eq:ss2c}
    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
where $\bm{\theta}$ are the process variation parameters governing
initial abundance, $\bm{\Theta}$ are the process variation parameters
influencing how abundance changes over time, and $\bm{p}$ are the
observation error parameters. In our first example, Eq.~\ref{eq:ss1},
But what should these distributions be, and how could
parameters of the distributions be
estimated since there is only a single observation available? In fact,
they cannotThe reason they were omitted is that it is not possible to
estimate the parameters of the i
However, a completSpecifically, the Markovian
nature of the model implies the following likelihood
The first term is the probability density for the first observation in
the time series, $X_1$, yet it is intuitively not possible to estimate the
parameter(s) of this distribution using only a single
observation.
%This problem has resulted in several papers with
%interesting names like ``multi-modal likelihoods and ridges etc...''
Two options have been proposed handling the identifiability
problem. First, one can assume that the first observation in the time
series is made without error. Second, one can assume that the population is at equilibrium
so that that the x can be replaced with the equilibrium
distribution. However, assuming equilibrium defeats the objective of
many studies of population dynamics, namely determining if and why a
population is at equilibrium.
